Teaching Low-Level Concurrency with \pebbles and \landslide

### 2016-08-27 sigcse-17/410.tex
sigcse: SAVE THE ANIMALS!^Wlines.

The course has many learning objectives,
ranging from acquiring detailed factual knowledge about
hardware features
through practicing advanced cognitive processes
such as open-ended design.
%%
%Students study high-level concepts
%such as
%\shortversion{least privilege}{protection (least privilege, access control lists vs.\ capabilities)},
%file-system internals,
%and log-based storage.
%%
%Emphasis is placed on acquiring information from ``primary sources,''
%including both manufacturer-provided hardware documentation
%and a non-textbook technical-literature reading assignment.
%
The course's distinguishing feature is that
students begin with a ``blank slate'' rather than a
kernel-source template or an existing operating system,
so they must synthesize design requirements from multiple sources
and must
choose their own module boundaries and inter-module conventions.
%
%Due to the foundational nature of kernel code,
The assignment design and grading encourage students to
think about corner cases, including resource exhaustion,
instead of being satisfied by
``the right basic idea''
implementations that handle only auspicious situations.
%
Finally, most relevant to this work,
students %gain substantial experience in
analyze and write lock-based multi-threaded code and
thread-synchronization objects.
They practice detecting and documenting deadlock and race conditions,
including both thread/thread
%concurrency
and thread/interrupt concurrency.

@@

\textbf{32-bit x86.}
%\subsubsection{x86}
%\shortversion{}{Students write code that runs on one specific
%hardware platform.}
We target x86 PCs because machines of this type are
available from many manufacturers at many price points.
This means that it is easy for students to obtain access
to a machine (virtual or real) which will let them
test their code.
%boot
%up their code, type commands, and observe the results.
%\violence{}{Also, unlike many System-on-a-Chip platforms,
%detailed hardware descriptions are readily available.}
%% rather than being proprietary or "documented"
%% by labyrinthine Linux source code
%\shortversion{}{Finally, the platform is fairly stable---with only a
%little care, it is possible to write a compliant
%kernel which runs on anything newer than a 486DX processor,
%i.e., almost any PC hardware which is likely to be in
%working condition.}
%
%\subsubsection{32-bit}
We target the 32-bit
subset of x86, for multiple reasons.
%%% I'm not sure I believe this is an advantage. Instead of having
%%% to verify if each register is correct, students now have to verify
%%% both registers and stack slots. Otherwise this paragraph is good. -- bblum
%First, while the small number of registers is
%frustrating for compiler writers,
%\shortversion{the}{from the point of view of OS students, the}
%existence of fewer registers means less work
%when it is necessary to verify that each register's
%value is correct.
%Likewise,
The calling convention's old-fashioned focus on the stack
makes it easier to track function parameter values.
Second, implementing virtual memory on x86-64 as
opposed to x86-32 requires more code and more
debugging time without imparting greater
insight.
%\shortversion{insight.}{insight. Finally, since it is impossible for an x86 processor
%to be in 64-bit mode with virtual memory off,
%students would need to be coached through the
%32-bit--to--64-bit transition or would need
%to be given a template virtual-memory implementation.}
%All in all, at present we are satisifed with
%the 32-bit choice.

>>>>

The course's learning objectives
range from acquiring detailed knowledge about
hardware features
to practicing advanced programming skills %cognitive processes
such as open-ended design.
%%
%Students study high-level concepts
%such as
%\shortversion{least privilege}{protection (least privilege, access control lists vs.\ capabilities)},
%file-system internals,
%and log-based storage.
%%
%Emphasis is placed on acquiring information from ``primary sources,''
%including both manufacturer-provided hardware documentation
%and a non-textbook technical-literature reading assignment.
%
%The course's distinguishing feature is that
Students begin with a blank slate rather than a
kernel-source template or an existing operating system,
so they must synthesize design requirements %from multiple sources
and %must
choose their own module boundaries and conventions.
%
%Due to the foundational nature of kernel code,
The assignment design and grading encourage students to
think about corner cases, including resource exhaustion,
instead of being satisfied by
``the right basic idea''
implementations that handle only auspicious situations.
%
Finally, most relevant to this work,
students %gain substantial experience in
analyze and write lock-based multi-threaded code and
thread-synchronization objects.
They practice detecting and documenting deadlock and race conditions,
including both thread/thread
%concurrency
and thread/interrupt concurrency.

@@

\textbf{32-bit x86.}
%\subsubsection{x86}
%\shortversion{}{Students write code that runs on one specific
%hardware platform.}
We target x86 PCs %because machines of this type are
%available from many manufacturers at many price points.
%This means that it is easy for
so students can easily obtain access
to a machine (virtual or real) which will let them
test their code.
%boot
%up their code, type commands, and observe the results.
%\violence{}{Also, unlike many System-on-a-Chip platforms,
%detailed hardware descriptions are readily available.}
%% rather than being proprietary or "documented"
%% by labyrinthine Linux source code
%\shortversion{}{Finally, the platform is fairly stable---with only a
%little care, it is possible to write a compliant
%kernel which runs on anything newer than a 486DX processor,
%i.e., almost any PC hardware which is likely to be in
%working condition.}
%
%\subsubsection{32-bit}
We target the 32-bit
subset of x86, for multiple reasons.
%%% I'm not sure I believe this is an advantage. Instead of having
%%% to verify if each register is correct, students now have to verify
%%% both registers and stack slots. Otherwise this paragraph is good. -- bblum
%First, while the small number of registers is
%frustrating for compiler writers,
%\shortversion{the}{from the point of view of OS students, the}
%existence of fewer registers means less work
%when it is necessary to verify that each register's
%value is correct.
%Likewise,
The calling convention's old-fashioned focus on the stack
makes it easier to track function parameter values.
Second, implementing virtual memory on x86-64 as
opposed to x86-32 requires more code and more
debugging time without imparting greater
insight.
%\shortversion{insight.}{insight. Finally, since it is impossible for an x86 processor
%to be in 64-bit mode with virtual memory off,
%students would need to be coached through the
%32-bit--to--64-bit transition or would need
%to be given a template virtual-memory implementation.}
%All in all, at present we are satisifed with
%the 32-bit choice.

### 2016-08-27 sigcse-17/eval.tex
sigcse: SAVE THE ANIMALS!^Wlines.

Whenever a student ran \landslide,
it recorded which options were used (test program and time limit),
a snapshot of the student's code,
and the result of the test (completed, timed out, bug found, or ctrl-C'ed).
%
We manually analyzed these snapshots to count how many bugs each group found
(not double-counting the same bug found repeatedly)
and how many of those were fixed before submission
(i.e., the students changed their code and re-ran the same test successfully).
%
We also distinguished {\em deterministic} bugs from concurrency bugs
by identifying whether \landslide needed to test multiple interleavings before finding the bug\footnote{
	For example, many students found deterministic use-after-free bugs,
	previously overlooked because the unit tests do not provide heap checking like \landslide does.
}.

@@

\begin{figure}[t]
	\includegraphics[width=0.48\textwidth]{p2-distribution-logos.pdf}
	\caption{Distribution of thread library grades.}
	\label{fig:p2-distribution}
\end{figure}
We also collected the overall thread library grades to evaluate whether \landslide helps students achieve more during the project.
Figure~\ref{fig:p2-distribution} shows the distribution of these grades.
Compared to students from the same semesters who opted not to use \landslide,
those who did scored on average 3\% higher. % that's rough, buddy.
%
%However, to account for the possible bias for opting-out students being more at the bottom of the class
%on account of not enough free time to volunteer for \landslide,
However, the latter distribution is potentially biased towards struggling students
who lack enough free time to volunteer for \landslide.
%To account for this,
Thus, we also checked grades from prior semesters when \landslide was not available.
%This distribution is indistiguishable from the experimental group,
Although the minimum grade among the experimental group is 17\% higher than the minimum from prior semesters,
the middle 50\% of the distributions are quite similar,
suggesting that \landslide's measurable benefit is largely offset by the time investment required.
%
%In future work, we will streamline the user experience to help students fix as many concurrency bugs with a smaller time investment.

>>>>

Whenever a student ran \landslide,
it recorded which options were used (test program and time limit),
a snapshot of the student's code,
and the result of the test (completed, timed out, bug found, or ctrl-C'ed).
%
We manually analyzed these snapshots to count how many bugs each group found
(not double-counting the same bug found repeatedly)
and how many of those were fixed before submission
(i.e., the students changed their code and re-ran the same test successfully).
%
We also distinguished {\em deterministic} bugs from concurrency bugs
by identifying whether \landslide needed to test multiple interleavings before finding the bug.
(For example, many students found deterministic use-after-free bugs,
previously overlooked because the unit tests do not provide heap checking like \landslide does.)

@@

\begin{figure}[t]
	\includegraphics[width=0.48\textwidth]{p2-distribution-logos-squashed.pdf}
	\caption{Distribution of thread library grades.}
	\label{fig:p2-distribution}
\end{figure}
We also collected the overall thread library grades to evaluate whether \landslide helps students achieve more during the project.
Figure~\ref{fig:p2-distribution} shows the distribution of these grades.
Compared to students from the same semesters who opted not to use \landslide,
those who did scored on average 3\% higher. % that's rough, buddy.
%
%However, to account for the possible bias for opting-out students being more at the bottom of the class
%on account of not enough free time to volunteer for \landslide,
However, the latter distribution is potentially biased towards struggling students
who lack enough free time to volunteer for \landslide.
%To account for this,
Thus, we also checked grades from prior semesters when \landslide was not available.
%This distribution is indistiguishable from the experimental group,
Although the minimum grade among the experimental group is 17\% higher than the minimum from prior semesters,
the middle 50\% of the distributions are quite similar,
suggesting that \landslide's measurable benefit is largely offset by the time investment required.
%
%In future work, we will streamline the user experience to help students fix as many concurrency bugs with a smaller time investment.

### 2016-08-27 sigcse-17/landslide.tex
sigcse: SAVE THE ANIMALS!^Wlines.

mutex_unlock(mx);
\end{lstlisting}
	\end{tabular}
	{\bf (a}) A buggy producer/consumer program, in which the consumer thread only checks \x{work_exists()} once.
	\vspace{1em}
	\\
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%	\begin{tabular}{l|l|l}
%	\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} & \cellcolor{thread3} {\bf Thread 3} \\
%	\hline
%\begin{lstlisting}
%lock();
%if (!work())
%  wait();
%\end{lstlisting}
%& & \\
%&
%\begin{lstlisting}
%lock();
%enqueue();
%signal();
%unlock();
%\end{lstlisting}
%& \\
%& &
%\begin{lstlisting}
%lock();
%dequeue();
%unlock();
%\end{lstlisting}
%\\
%\begin{lstlisting}
%dequeue();
%unlock();
%// SIGSEGV
%\end{lstlisting}
%& &
%	\begin{center}
%	{\bf (b)} Interleaving with two consumer threads. The second bypasses \x{cond_wait()}, causing the first to receive \x{NULL}. %in which one receives \x{NULL}.
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	\includegraphics[width=0.48\textwidth]{execution-tree.pdf}
	{\bf (b)} The state space of all interleavings forms an execution tree. %The interleaving in (b) is the rightmost branch.
	The bug is observed when the second consumer thread bypasses \x{cond_wait()}, causing the first to receive \x{NULL}.
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	\caption{Example concurrency bug.}
	\label{fig:paradise}
\end{figure}

@@

\landslide checks for bugs in several ways.
It can identify assertion failures, use-after-free and other invalid heap accesses, and
deadlocks, and also includes a heuristic check for infinite loops and livelock.
%Note that our use of the term ``race condition'' throughout this paper refers to concurrency errors in general, a category that includes data races, atomicity violations, and nondeterministic deadlocks.
%In contrast with analyses that directly detect data races~\cite{tsan},
Although \landslide employs a data-race analysis to decide where to preempt,
it does not directly classify suspicious memory accesses as bugs,
but rather seeks concrete, observable failures that can arise from many different types of errors.

@@

>>>>

mutex_unlock(mx);
\end{lstlisting}
	\end{tabular}
	{\bf (a}) A buggy producer/consumer program, in which the consumer thread only checks \x{work_exists()} once.
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%	\begin{tabular}{l|l|l}
%	\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} & \cellcolor{thread3} {\bf Thread 3} \\
%	\hline
%\begin{lstlisting}
%lock();
%if (!work())
%  wait();
%\end{lstlisting}
%& & \\
%&
%\begin{lstlisting}
%lock();
%enqueue();
%signal();
%unlock();
%\end{lstlisting}
%& \\
%& &
%\begin{lstlisting}
%lock();
%dequeue();
%unlock();
%\end{lstlisting}
%\\
%\begin{lstlisting}
%dequeue();
%unlock();
%// SIGSEGV
%\end{lstlisting}
%& &
%	\begin{center}
%	{\bf (b)} Interleaving with two consumer threads. The second bypasses \x{cond_wait()}, causing the first to receive \x{NULL}. %in which one receives \x{NULL}.
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	\includegraphics[width=0.48\textwidth]{execution-tree.pdf}
	{\bf (b)} The state space of all interleavings forms an execution tree. %The interleaving in (b) is the rightmost branch.
	The bug is observed when the second consumer thread bypasses \x{cond_wait()}, causing the first to receive \x{NULL}.
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	\caption{Example concurrency bug.}
	\label{fig:paradise}
\end{figure}

@@

@@

\landslide %checks for bugs in several ways.
%It can identify
identifies assertion failures, use-after-free and other invalid heap accesses, and
deadlocks, and also includes a heuristic check for infinite loops and livelock.
%Note that our use of the term ``race condition'' throughout this paper refers to concurrency errors in general, a category that includes data races, atomicity violations, and nondeterministic deadlocks.
%In contrast with analyses that directly detect data races~\cite{tsan},
Although \landslide employs a data-race analysis to decide where to preempt,
it does not directly classify suspicious memory accesses as bugs,
but rather seeks concrete, observable failures that can arise from many different types of errors.
